Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-06-01 paper/AlignedIsotropicGW.tex
Consistency in 'Methods Section XX'.  Close #39

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``low'' magnitude distribution; the most favoured
model among those with an aligned distribution also has the ``low''
magnitude distribution.  The odds ratio between the ``low'' aligned
and ``low'' isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
\ac{BBH} systems having zero spin is ruled out by the GW151226
measurements, which bound at least one black hole to have spin
magnitude $\geq 0.2$ at 99\% credibility\cite{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two
years\cite{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming
10 additional detections drawn from each of our six zero-parameter
models for the spin distribution in addition to the three existing
detections from O1, with observational uncertainties drawn randomly
from the three Gaussian widths used to approximate the $\chieff$
posteriors in Figure \ref{fig:O1-posteriors}\footnote{The measurement
  uncertainty in $\chieff$ depends on the other parameters of the
  merging \ac{BBH} system, particularly on the mass ratio.  Our
  assumption about future observational uncertainties is appropriate
  if the parameters of the three detected events are representative of
  the parameters of future detections. See Section
  \ref{sec:chi-eff-precision} for further discussion.}, we find the
odds ratios shown in Figure \ref{fig:O2-predictions}.  We find that
most scenarios with an additional 10 detections allow the simulated
angular distribution to be inferred with greater than $5\sigma$
($2.9 \times 10^{-7}$ odds) credibility; and in the most pessimistic
case the distinction is typically $\OTwoSigmaIsoAlignedMin\sigma$
($\OTwoOddsIsoAlignedMin$ odds ratio).  While such future detections
should permit a confident distinction between \emph{angular}
distributions, we would remain much less certain about the
\emph{magnitude} distribution among the three options considered here
until we have a larger number of
observations\cite{2017arXiv170306869F,2017arXiv170306223G}.  In Figure
\ref{fig:O2-predictions}, the odds ratio between different magnitude
distributions with the same angular distribution is much closer to
unity than the odds ratio between angular distributions.

@@

If the true distribution is isotropic and follows this simple model,
and our measurement uncertainties on $\chieff$ are $\simeq 0.1$, then
the geometric mean of each subsequent measurement's contribution to
the overall odds is $\sim 3$.  After ten additional events, then, the
odds ratio becomes $5 \times 3^{10} \simeq 3 \times 10^{5}$, or
$4.6 \sigma$, consistent with the results of the more detailed
calculation described above.  If the true distribution of spins
becomes half as wide ($\chieff \in [-0.125, 0.125]$ for isotropic and
$\chieff \in [0, 0.25]$ for aligned spins), with the same
uncertainties, then the existing odds ratio becomes $1.08$, and each
subsequent event drawn from the isotropic distribution contributes on
average a factor of $1.6$.  In this case, after 10 additional events,
the odds ratio becomes $150$, or $2.7\sigma$.  With small spin
magnitudes, our angular resolving power vanishes, as discussed in more
detail in Section \ref{sec:smallspins}.

>>>>

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Methods Section \ref{sec:hierarchical}).  We also fit three mixture
models for the population, where the spin magnitude distribution is
fixed but the angular distribution is a weighted sum of the isotropic
and aligned distribution.  The evidence, or marginal likelihood, for
each of the models is shown in Figure \ref{fig:O1-odds}.  For all
three magnitude distributions, the mixture models' posterior on the
mixing fraction peaks at 100\% isotropic, which explains why the
zero-parameter, pure-isotropic models are preferred over the
single-parameter mixture models for every magnitude distribution with
this data set.  Not surprisingly, given the small $\chieff$ values in
the three detected systems, the most-favoured model among those with
an isotropic angular distribution has the ``low'' magnitude
distribution; the most favoured model among those with an aligned
distribution also has the ``low'' magnitude distribution.  The odds
ratio between the ``low'' aligned and ``low'' isotropic models is
$\OOneOddsIsoAligned$, or $\OOneSigmaIsoAligned\sigma$; thus the data
favour isotropic spins among our suite of models.  While the data
favour spin amplitude distributions with small spin magnitudes, note
that a model with all \ac{BBH} systems having zero spin is ruled out
by the GW151226 measurements, which bound at least one black hole to
have spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two
years\cite{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming
10 additional detections drawn from each of our six zero-parameter
models for the spin distribution in addition to the three existing
detections from O1, with observational uncertainties drawn randomly
from the three Gaussian widths used to approximate the $\chieff$
posteriors in Figure \ref{fig:O1-posteriors}\footnote{The measurement
  uncertainty in $\chieff$ depends on the other parameters of the
  merging \ac{BBH} system, particularly on the mass ratio.  Our
  assumption about future observational uncertainties is appropriate
  if the parameters of the three detected events are representative of
  the parameters of future detections. See Methods Section
  \ref{sec:chi-eff-precision} for further discussion.}, we find the
odds ratios shown in Figure \ref{fig:O2-predictions}.  We find that
most scenarios with an additional 10 detections allow the simulated
angular distribution to be inferred with greater than $5\sigma$
($2.9 \times 10^{-7}$ odds) credibility; and in the most pessimistic
case the distinction is typically $\OTwoSigmaIsoAlignedMin\sigma$
($\OTwoOddsIsoAlignedMin$ odds ratio).  While such future detections
should permit a confident distinction between \emph{angular}
distributions, we would remain much less certain about the
\emph{magnitude} distribution among the three options considered here
until we have a larger number of
observations\cite{2017arXiv170306869F,2017arXiv170306223G}.  In Figure
\ref{fig:O2-predictions}, the odds ratio between different magnitude
distributions with the same angular distribution is much closer to
unity than the odds ratio between angular distributions.

@@

If the true distribution is isotropic and follows this simple model,
and our measurement uncertainties on $\chieff$ are $\simeq 0.1$, then
the geometric mean of each subsequent measurement's contribution to
the overall odds is $\sim 3$.  After ten additional events, then, the
odds ratio becomes $5 \times 3^{10} \simeq 3 \times 10^{5}$, or
$4.6 \sigma$, consistent with the results of the more detailed
calculation described above.  If the true distribution of spins
becomes half as wide ($\chieff \in [-0.125, 0.125]$ for isotropic and
$\chieff \in [0, 0.25]$ for aligned spins), with the same
uncertainties, then the existing odds ratio becomes $1.08$, and each
subsequent event drawn from the isotropic distribution contributes on
average a factor of $1.6$.  In this case, after 10 additional events,
the odds ratio becomes $150$, or $2.7\sigma$.  With small spin
magnitudes, our angular resolving power vanishes, as discussed in more
detail in Methods Section \ref{sec:smallspins}.
